Presheaf Model for Dependent Types

### 2017-09-21 papers/talk-11/paper-11.tex
Update the writeup11

\begin{example}
  We always have the empty context morphism $()$ from any context $\Gamma$ to
  $\diamond$, it does not contain any terms and is the only context morphism
  from $\Gamma$ to $\diamond$.
\end{example}
\begin{example}
  We also have the identity context morphism $1$ (from now on we will use $1$
  as the suitable identity morphism for the surroundings) for any $\Gamma = x_1
  : A_1, \dots, x_n : A_n$, which is $(x_1, \dots, x_n)$.
\end{example}

A note on notation: we sometimes write $A\sigma$, where $\Gamma \vdash
A\sigma~type$, $\Delta \vdash A~type$, and $\sigma \from \Gamma \to \Delta$ for
some contexts $\Gamma$ and $\Delta$, to denote the application of $\sigma$ on
the right. That is, $A\sigma$ becomes here a type in the domain context of
$\sigma$, where $\sigma(A\sigma) = A$. The same notation is sometimes
analogously used for terms.

\begin{definition}[Context extension]
  If $\vdash \Gamma~ctxt$ and $\Gamma \vdash A~type$, there is a context
  $\Gamma, x:A$ (for brevity also written as $\Gamma.A$ from now on), a context
  morphism $\mathbf{p}: \Gamma.A \to \Gamma$, and a term $\Gamma.A \vdash
  \mathbf{q} : A \mathbf{p}$. This should satisfy the following property: for
  each $\vdash \Delta~ctxt$, a context morphism $\sigma \from \Delta \to
  \Gamma$, and $\Delta \vdash a : A \sigma$, there is a context morphism
  $(\sigma, a): \Delta \to \Gamma.A$ called \emph{context extension} such that:
  \[
    \mathbf{p}(\sigma, a) = \sigma \quad \mathbf{q}(\sigma, a) = a \quad (\sigma, a) \tau =
    (\sigma \tau, a \tau) \quad (\mathbf{p}, \mathbf{q}) = 1
  \]
  (Note that in the last equation $1: \Gamma.A \to \Gamma.A$)
\end{definition}

@@

  \item[$\bullet$] $F$ is a functor $F\from \cat{C}^\op \to \fcat{Fam}$; For a
    context $\Gamma$ we write $Ty_F(\Gamma)$ for the indexing set of the family
    $F(\Gamma)$ and its family as $(Ter_F(\Gamma;A) | A \in Ty_F(\Gamma))$;
    $\Gamma \vdash A~type$ means that $A \in Ty_F(\Gamma)$ and $\Gamma \vdash a
    : A$ means $a \in Ter_F(\Gamma;A)$. For a context morphism $\sigma \from
    \Delta \to \Gamma$ the morphism $F(\sigma)$ acts on types $\Gamma \vdash
    A~type$ as $A\sigma$ and on terms $\Gamma \vdash a : A$ as $a\sigma$. The
    following equations also hold:
    \[
      A1=A \quad (A\sigma)\tau = A(\sigma\tau) \quad a1=a \quad (a\sigma)\tau=a(\sigma\tau)
    \]
\end{itemize}
\end{definition}

We say that a category with families supports $\Pi$- and $\Sigma$-types if it is
closed under the corresponding rules shown in previous sections.

We also add the rules for context morphisms:

@@

We will now show how any presheaf category gives rise to a category with
families where the contexts are presheaves.  Let $\cat{C}$ be a small category.
In the following objects of $\cat{C}$ will be denoted as $I,J,K$ and morphisms
as $f,g,h$.

@@

The empty context then is a terminal presheaf and a context morphism $\sigma
\from \Delta \to \Gamma$ is a natural transformation making the diagram
commute.

\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]

If we write $\Gamma_f$ and $\Delta_f$ as $f$ acting on the right, the
equation becomes:
\[
  (\sigma \rho) f = \sigma (\rho f)
\]
for $\rho$ in $\Delta(I)$.

>>>>

\begin{example}
  We always have the empty context morphism $()$ from any context $\Gamma$ to
  $\diamond$, it does not contain any terms and is the only context morphism
  from $\Gamma$ to $\diamond$.
\end{example}
\begin{example}
  We also have the identity context morphism $1$ for any $\Gamma = x_1 : A_1,
  \dots, x_n : A_n$, which is $(x_1, \dots, x_n)$.
\end{example}

A note on notation: we sometimes write $A\sigma$, where $\Gamma \vdash
A\sigma~type$, $\Delta \vdash A~type$, and $\sigma \from \Gamma \to \Delta$ for
some contexts $\Gamma$ and $\Delta$, to denote the application of $\sigma$ on
the right. That is, $A\sigma$ becomes here a type in the domain context of
$\sigma$, where $\sigma(A\sigma) = A$. The same notation is sometimes
analogously used for terms: $a\sigma$ to denote a term $\Gamma \vdash a\sigma :
A\sigma$, where $\Delta \vdash a : A$.

Having this notation, we also want the following equations to hold:
\[
  A1=A \quad (A\sigma)\tau = A(\sigma\tau) \quad a1=a \quad (a\sigma)\tau=a(\sigma\tau)
\]
For readability, we usually omit writing indices to the identity morphism $1$,
using it as the most suitable identity morphism for the surroundings. We also
assume correct types in the equations, i.e. $(A\sigma)\tau = A(\sigma\tau)$
assumes $\sigma \from \Delta \to \Gamma$, $\tau \from \Theta \to \Delta$, and
$\Gamma \vdash A~type$.

\begin{definition}[Context extension]
  If $\vdash \Gamma~ctxt$ and $\Gamma \vdash A~type$, there is a context
  $\Gamma, x:A$ (for brevity also written as $\Gamma.A$ from now on), a context
  morphism $\mathbf{p}: \Gamma.A \to \Gamma$, and a term $\Gamma.A \vdash
  \mathbf{q} : A \mathbf{p}$. This should satisfy the following property: for
  each $\vdash \Delta~ctxt$, a context morphism $\sigma \from \Delta \to
  \Gamma$, and $\Delta \vdash a : A \sigma$, there is a context morphism
  $(\sigma, a): \Delta \to \Gamma.A$ called \emph{context extension} such that:
  \[
    \mathbf{p}(\sigma, a) = \sigma \quad \mathbf{q}(\sigma, a) = a \quad (\sigma, a) \tau =
    (\sigma \tau, a \tau) \quad (\mathbf{p}, \mathbf{q}) = 1
  \]
  Note that in the last equation $1: \Gamma.A \to \Gamma.A$ and that we leave
  the indices on $\mathbf{p}$ implicit.
\end{definition}

@@

  \item[$\bullet$] $F$ is a functor $F\from \cat{C}^\op \to \fcat{Fam}$; For a
    context $\Gamma$ we write $Ty_F(\Gamma)$ for the indexing set of the family
    $F(\Gamma)$ and its family as $(Ter_F(\Gamma;A)~|~A \in Ty_F(\Gamma))$;
    $\Gamma \vdash A~type$ means that $A \in Ty_F(\Gamma)$ and $\Gamma \vdash a
    : A$ means $a \in Ter_F(\Gamma;A)$. Note that for a context morphism
    $\sigma \from \Delta \to \Gamma$ the morphism $F(\sigma)$ acts on types
    $\Gamma \vdash A~type$ as $A\sigma$ and on terms $\Gamma \vdash a : A$ as
    $a\sigma$.
\end{itemize}
\end{definition}

A category with families supports $\Pi$- and $\Sigma$-types if it is
closed under the $\Pi$ and $\Sigma$ rules.

@@

Let us now consider the categories of presheaves and see that they form a
category with families where the contexts are presheaves.  Let $\cat{C}$ be a
small category. We will refer to the objects of $\cat{C}$  as $I,J,K$ and
morphisms as $f,g,h$.

@@

The empty context $\diamond$ is a terminal presheaf and a context morphism
$\sigma \from \Delta \to \Gamma$ is a natural transformation that makes the
diagram commute.
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]

In the following, we will omit the indices on $\sigma$; Then if we write
$\Gamma_f$ and $\Delta_f$ as $f$ acting on the right, the equation becomes:
\[
  (\sigma \rho) f = \sigma (\rho f)
\]
for $\rho$ in $\Delta(I)$.
